---
title: "ML - Probability Practice"
author: "Sarah Dominguez"
date: "2024-08-12"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Part A.** Visitors to your website are asked to answer a single survey
question before they get access to the content on the page. Among all of
the users, there are two categories: Random Clicker (RC), and Truthful
Clicker (TC). There are two possible answers to the survey: yes and no.
Random clickers would click either one with equal probability. You are
also giving the information that the expected fraction of random
clickers is 0.3. After a trial period, you get the following survey
results: 65% said Yes and 35% said No. What fraction of people who are
truthful clickers answered yes? Hint: use the rule of total probability.

## **Solution:**

**Step 1:** **Define Initial Probabilities**

**Random Clicker (RC):** Clicks "Yes" or "No" with equal probability:

-   *P(Yes\|RC) = 0.5*

-   *P(No\|RC) = 0.5*

**Truthful Clickers (TC)**: Someone who always clicks the correct
answer:

-   **Probability of a user being a Random Clicker (RC):**

    *P(RC)=0.3*

-   **Probability of a user being a Truthful Clicker (TC):**

    *P(TC) = 1-P(RC) = 0.7*

**Step 2:** **Survey Results After the trial period, the survey results
indicate:**

-   *P(Yes)=0.65 - P(No)=0.35*

**Step 3:** **Calculate the Fraction of Truthful Clickers Who Answered
"Yes" Using the Rule of Total Probability:**

-   *P(Yes)= P(Yes∣TC) x P(TC) + P(Yes\|RC) x P(RC)*

```{r Probability of Truthful Clickers that answer yes}
# Calculate the probability of a truthful clicker answering "Yes"
prob_yes_truth <- (0.65 - (0.5 * 0.3)) / 0.7

# Convert to percentage
fraction <- prob_yes_truth * 100

# Print the result
print(paste("The fraction of people who are truthful clickers answered 'Yes' is", round(fraction, 2), "%"))
```

**Part B**

Imagine a medical test for a disease with the following two attributes:

-   The sensitivity is about 0.993. That is, if someone has the disease,
    there is a probability of 0.993 that they will test positive.
-   The specificity is about 0.9999. This means that if someone doesn't
    have the disease, there is probability of 0.9999 that they will test
    negative.
-   In the general population, incidence of the disease is reasonably
    rare: about 0.0025% of all people have it (or 0.000025 as a decimal
    probability).

Suppose someone tests positive. What is the probability that they have
the disease?

## **Solution:**

**Step 1: Define Initial Probabilities:**

-   *P(D) = 0.000025 (Probability of having the disease)*

-   *P(NoD) = 1 − 0.000025 = 0.999975 (Probability of not having the
    disease)*

**Sensitivity of the test is 0.993, leading to:**

-   *P(+Pos∣D) = 0.993 (Probability of testing positive given that the
    individual has the disease)*

-   *P(-Neg∣D) = 1 − 0.993 = 0.007 (Probability of testing negative
    given that the individual has the disease)*

**Specificity of the test is 0.9999, leading to:**

-   *P(-Neg∣NoD) = 0.9999 (Probability of testing negative given that
    the individual does not have the disease)*

-   *P(+Pos∣NoD) = 1−0.9999 = 0.0001 (Probability of testing positive
    given that the individual does not have the disease)*

### **Question: "Suppose someone tests positive. What is the probability that they actually have the disease?"**

We calculate the following joint probabilities using Bayes' Theorem:

-   P(+Pos 'and' D) = P(+Pos∣D) × P(D)

-   P(+Pos 'and' NoD) = P(+Pos∣NoD) × P(NoD)

```{r}
# Calculating joint probabilities
joint_prob_pos_disease <- 0.993 * 0.000025
joint_prob_pos_no_disease <- 0.0001 * 0.9999

# Calculate the total probability of testing positive
total_prob_positive <- joint_prob_pos_disease + joint_prob_pos_no_disease
prob_disease_given_pos <- joint_prob_pos_disease / total_prob_positive
prob_no_disease_given_pos <- joint_prob_pos_no_disease / total_prob_positive

# Display the result
prob_disease_given_pos
print(paste("The probability of someone having the disease given that they tested positive is", round(prob_disease_given_pos * 100, 2), "%"))

```
