---
title: 'Visual Story Telling Part 2: Capital Metro Data'
author: "Sarah Dominguez"
date: "2024-08-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Visual story telling part 2: Capital Metro data

The file [capmetro_UT.csv](https://github.com/jgscott/STA380/blob/master/data/capmetro_UT.csv) contains data from Austin's own Capital Metro bus network, including shuttles to, from, and around the UT campus. These data track ridership on buses in the UT area. Ridership is measured by an optical scanner that counts how many people embark and alight the bus at each stop. Each row in the data set corresponds to a 15-minute period between the hours of 6 AM and 10 PM, each and every day, from September through November 2018. The variables are:

-   *timestamp*: the beginning of the 15-minute window for that row of data

-   *boarding*: how many people got on board any Capital Metro bus on the UT campus in the specific 15 minute window

-   *alighting*: how many people got off ("alit") any Capital Metro bus on the UT campus in the specific 15 minute window

-   *day_of_week* and *weekend*: Monday, Tuesday, etc, as well as an indicator for whether it's a weekend.

-   *temperature*: temperature at that time in degrees F

-   *hour_of_day*: on 24-hour time, so 6 for 6 AM, 13 for 1 PM, 14 for 2 PM, etc.

-   *month*: July through December

Your task is to create a figure, or set of related figures, that tell an interesting story about Capital Metro ridership patterns around the UT-Austin campus during the semester in question. Provide a clear annotation/caption for each figure, but the figure(s) should be more or less stand-alone, in that you shouldn't need many, many paragraphs to convey its meaning. Rather, the figure together with a concise caption should speak for itself as far as possible.

You have broad freedom to look at any variables you'd like here -- try to find that sweet spot where you're showing genuinely interesting relationships among more than just two variables, but where the resulting figure or set of figures doesn't become overwhelming/confusing. (Faceting/panel plots might be especially useful here.)

```{r read in the data}
library(ggplot2)
library(dplyr)
library(readr)

# Load the data
capmetro <- read.csv("capmetro_UT.csv")

# View the first few rows of the data
head(capmetro)

```

```{r Ridership by the time of the day and day of the week (weekdays)}

# Splitting the data into weekday and weekend
weekday_data <- capmetro %>% filter(weekend == "weekday")
weekend_data <- capmetro %>% filter(weekend == "weekend")


ggplot(weekday_data, aes(x = hour_of_day, y = boarding + alighting, color = day_of_week)) +
  geom_line(stat = "summary", fun = "mean") +
  scale_x_continuous(breaks = seq(6, 22, by = 1)) +  # Display all hours
  labs(title = "Average Ridership by Time of Day (Weekdays)",
       x = "Hour of Day",
       y = "Average Number of Riders",
       color = "Day of Week",
       caption = "This chart illustrates the ridership patterns for Capital Metro buses from Monday to Friday,\nhighlighting distinct morning and evening peaks corresponding to typical commuter rush hours.\nThe morning peak is the most pronounced, occurring between 7 AM and 9 AM, while\nthe evening peak occurs from 4 PM to 6 PM. There is a slight variation in ridership on Fridays,\nparticularly in the evening, which could reflect different\ncommuter behaviors as the weekend approaches.") +
  theme_minimal()
```

```{r}
ggplot(weekend_data, aes(x = hour_of_day, y = boarding + alighting, color = day_of_week)) +
  geom_line(stat = "summary", fun = "mean") +
  scale_x_continuous(breaks = seq(6, 22, by = 1)) +  # Display all hours
  labs(title = "Average Ridership by Time of Day (Weekends)",
       x = "Hour of Day",
       y = "Average Number of Riders",
       color = "Day of Week",
       caption = "This chart illustrates the weekend ridership patterns for Capital Metro\nbuses, showing a gradual increase in the morning and a stable ridership pattern during the\nmidday hours. Saturday exhibits higher ridership, especially during the evening, compared to Sunday.\nThe peak ridership on both days occurs between 4 PM and 6 PM, after which ridership declines sharply.") +
  theme_minimal()
```

```{r Ridership Distribution Across Temperature Ranges}
capmetro_clean2 <- capmetro %>%
  mutate(temperature = cut(temperature, breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 110)))

capmetro_clean2 <- capmetro_clean2 %>%
  filter(!is.na(temperature))

ggplot(capmetro_clean2, aes(x = temperature, y = boarding + alighting, fill = weekend)) +
  geom_boxplot() +
  labs(title = "Ridership Distribution Across Temperature Ranges",
       x = "Temperature Range (F)",
       y = "Number of Riders",
       fill = "Type of Day", 
       caption = "This boxplot illustrates the ridership patterns for Capital Metro buses across different\ntemperature ranges on weekdays and weekends.Weekday ridership is consistently\n higher than weekend ridership across all temperatures. The difference is especially\n noticeable in moderate temperatures (50°F - 70°F). As temperatures rise, ridership\n increases slightly for both weekdays and weekends, but weekdays still see more riders.\n Some days have unusually high ridership, especially when it’s warmer.") +
  theme_minimal()

```

```{r Cluster Analysis of Ridership Patterns}

library(dplyr)
library(ggplot2)
library(cluster)  
library(factoextra)  

# Combine 'boarding' and 'alighting' into a single 'ridership' metric
capmetro <- capmetro %>%
  mutate(ridership = boarding + alighting)

# Select relevant variables and scale them for clustering
clustering_data <- capmetro %>%
  select(hour_of_day, day_of_week, temperature, month, ridership) %>%
  mutate(day_of_week = as.numeric(as.factor(day_of_week)),  # Convert categorical to numeric
         month = as.numeric(as.factor(month))) %>%  # Convert categorical to numeric
  scale()  # Standardize the data

# Elbow method to find the optimal number of clusters
fviz_nbclust(clustering_data, kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal Clusters",
       caption = "Choosing the Right Number of Clusters: This graph shows how the variation within clusters\ndecreases as the number of clusters increases. The ideal number of clusters is typically at the 'elbow'\nof the curve, where adding more clusters doesn’t significantly reduce variation. In this case, the\nelbow seems to be around 6 clusters, suggesting that 6 clusters might be a good choice for grouping\nthe data effectively.")

set.seed(123)

# Perform k-means clustering with the best elbow
kmeans_result <- kmeans(clustering_data, centers = 6, nstart = 25)

# Add cluster to the original data
capmetro <- capmetro %>%
  mutate(cluster = as.factor(kmeans_result$cluster))

# Visualize clusters (hour_of_day vs. ridership)
ggplot(capmetro, aes(x = hour_of_day, y = ridership, color = cluster)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ day_of_week) +
  labs(title = "Cluster Analysis of Ridership Patterns",
       x = "Hour of Day",
       y = "Ridership",
       color = "Cluster", 
       caption = "This graph visualizes the results of a k-means clustering analysis, identifying six\ndistinct ridership patterns across different days of the week.\nWeekdays show clear morning and evening rush hour peaks, with specific clusters\ndominating during these times, indicating high and consistent ridership.\nIn contrast, weekends display more varied ridership patterns, with\nlower overall ridership and more spread-out usage throughout the day.") +
  theme_minimal()

# Print the cluster centers
print(kmeans_result$centers)

# Summary of the number of points in each cluster
table(capmetro$cluster)

```

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# turn data into by day of the week and hour
capmetro_hourly <- capmetro %>%
  mutate(hour = hour(timestamp)) %>%
  group_by(day_of_week, hour) %>%
  summarise(
    total_boarding = mean(boarding),
    total_alighting = mean(alighting)
  ) %>%
  ungroup() %>%
  arrange(day_of_week, hour)

# Create a line plot to visualize boarding vs. alighting over time
ggplot(capmetro_hourly, aes(x = hour)) +
  geom_line(aes(y = total_boarding, color = "Boarding"), size = 1) +
  geom_line(aes(y = total_alighting, color = "Alighting"), size = 1, linetype = "dashed") +
  facet_wrap(~ day_of_week, ncol = 2) +
  labs(title = "Average Passenger Flow of Boarding and Unboarding Throughout the Day",
       x = "Hour of Day",
       y = "Number of Passengers",
       color = "Flow Type",
       caption = "These plots show the average number of passengers boarding and alighting buses at\ndifferent times of the day for each day of the week. Weekdays (Monday to Friday) exhibit\nclear peaks during morning (8-9 AM) and afternoon (4-6 PM) commute times, with boarding\nand alighting patterns closely mirroring each other. On weekends (Saturday and Sunday), ridership\nis lower and more stable throughout the day. These trends can guide bus scheduling\nto better match passenger demand, especially during peak hours on weekdays.") +
  theme_minimal()


```
